Metric Spaces with Expensive Distances

### 2018-12-06 paper/draft.tex
Last touch

\pgfplotsset{compat=1.15}

@@

However, we argue that in some situations, distance computations
in $\metricspace$ can be costly and $\complexity$ might be incomparable
with $n$. Our motivation comes from topological summaries
such as persistence diagrams or Reeb graphs, which are of interest
in the field of topological data analysis. A persistence diagram
is a point set in $\R^2$, and the distance between two diagrams
is determined by a min-cost matching between the point sets.
If the diagrams have $N$ points, computing this matching requires
polynomial time in $N$, and $N$ might well be larger than $n$, the number
of diagrams considered (\cite{cohen2007stability}). For the case of Reeb graphs, the situation is even
worse: while several metrics on Reeb graphs have been proposed (\cite{bauer2014measuring}, \cite{de2016categorified},
\cite{di2016edit}),
not even an constant-factor approximation algorithm is known that runs
in polynomial time in the size of the graphs.
Another instance is a collection of high-resolution images
endowed with the Wasserstein (or Earth Movers) metric (\cite{rubner2000earth}).

@@

\myparagraph{Well-separated pair decomposition}
Given $t > 1$, two disjoint subsets $A, B$ of a metric space $(\metricspace, \dist)$ are called $t$-\textit{well-separated},
if 
\[
\forall a \in A \,\, \forall b \in B \,\, \dist(a, b) \geq t \max(\diam(A), \diam(B))
\]
A well-separated pair decomposition (WSPD) is a set of unordered pairs of sets $\{ \{A_1, B_1 \}, 
\dots, \{A_s, B_s\} \}$ such that each pair $\{A_i, B_i\}$ is $s$-well-separated, and for every unordered pair $\{a, b\}$ of distinct points of $\metricspace$ there exists a unique $j$ such that $a \in A_j$ and $b \in B_j$.
The notion of WSPD was introduced by Callahan and Kosaraju \cite{cal-kos-wspd} for Euclidean spaces.
Har-Peled and Mendel \cite{hm-fast} 
introduced the notion of net-trees and
generalized the results of \cite{cal-kos-wspd} for WSPD, proving the following:
\begin{enumerate}
    \item A net-tree for a metric space with $n$ points can be constructed in $2^{O(\mbox{dim})} n \log n)$ expected time.
    \item If $\{ \{A_1, B_1 \}, 
\dots, \{A_s, B_s\} \}$ is an $\eps / 16$-WSPD on $\metricspace$, and $a_i \in A_i, b_i \in B_i$ for $i = 1\dots s$
are chosen arbitrarily, then we get an $\eps$-spanner by taking $s$ edges $(a_i, b_i)$.
    \item For $\eps \in (0, 1]$,  an $\eps$-WSPD of size $n \eps^{-O({\dim})}$ can be constructed in $2^{O({\dim})} n \log n + n \eps ^{-O({\dim})}$ expected time. The algorithm uses the net-tree structure.
\end{enumerate}
The algorithm of constructing a net-tree is complicated and not easy to implement. Beygelzimer
et al. \cite{cover-trees} introduced the notion of a cover tree, which is a simpler data structure 
than net-trees. We mention in passing that cover trees can also be used for building
a spanner (this can be proven with the same methods), and we use cover trees for
building WSPD spanners in one of our implementations.

@@

\textsc{BlindQuasiSortedGreedy} tries to mimic the blind greedy spanner and hence appears as a natural
choice at first sight. However, anticipating the experimental results, the heuristic yields very poor
results. The reason is that no pair acquires useful lower bounds when only short distance are queried
(the greedy spanner does not have this issue because it knows the distance and hence does not need
lower bounds).
Generally speaking, short distances are good for sharp upper bounds, whereas long distances are
useful for lower bounds. This motivates \textsc{BlindQuasiSortedShaker}
which alternates between short and long distances.

@@

\section{Experiments on spanners}
We run experiments on the points sampled from the low-dimensional Euclidean space to investigate
experimentally the performance of these heuristics. Clearly, this has no practical meaning for $\R^d$, but
this space is well-suited for controlled experiments.
In order to test the \textsc{BlindQuasiSorted} algorithms we multiply the true distance by a factor from $[1,2]$ chosen uniformly.

@@

Figure \ref{fig:spanner_sparseness}
shows the number of edges of the spanner.
We can see that, even though none of the blind spanners can produce
spanners of the same quality (i.e., sparse) as the standard greedy algorithm,
\textsc{BlindGreedy} and all variants of \textsc{BlindRandom}
perform significantly better than both variants of \textsc{BlindQuasiSorted}.
It is important to remember that for all blind spanners the number of edges
is equal to the number of distances that were actually computed, while
for non-blind greedy spanner the latter number is always $\binom{n}{2}$.
Figure \ref{fig:spanner_ratio} shows the ratio of the number of edges to the number of points.
The ideal behavior is demonstrated by the non-blind greedy spanner,
for which this ratio stays practically constant, confirming the linear growth.
None of the blind algorithms seems to have this property, but among them
the blind greedy spanner is the best one. If we assume that the number of edges
is proportional to $n^\alpha$, then we can try to estimate $\alpha$ by 
linear regression (after taking $\log$).  We give in the table \ref{tbl:regr_coeff_spanner}
the estimated exponents $\alpha$ for \textsc{BlindGreedy} and standard
greedy algorithms. Note that even for the greedy algorithm these estimated
exponents can be significantly larger than 1,
which is explained by the fact that the number of points
on which we computed spanners is not large enough to clearly see
the linear dependence.

@@

We must also comment on WSPD spanners. In our experiments they performed poorly
on non-clustered data, while the plots in the extensive experimental study \cite{farshi2009experimental}
show that WSPD spanners are very sparse, outperformed only by the greedy algorithm.
We implemented two versions of  WSPD: one for the Euclidean case,
using quadtrees and the algorithm from \cite{hp-book}, and WSPD for general metric
spaces with cover trees, using the base 1.3, which is known to give much better practical
performance than the base 2. They both give similar results, and we can only conclude
that the advantage of WSPD shows up on larger point sets than the ones we deal with.
The paper \cite{farshi2009experimental} contains experiments for up to 30000 points,
and our blind algorithms, which have at least cubic complexity in the number of points,
are infeasible for such $n$.

We also look at higher dimensions, but the situation there is more
disappointing. We show the results for the best algorithm, \textsc{BlindGreedy},
in dimensions 2, 3 and 4 in the plot \ref{fig:blind_greedy_only}.
We can see that already in dimension 4 the best available algorithm
for the uniformly distributed point sets the blind greedy variant
produces almost a complete graph up to 700 points.

@@

Summing up, we can conclude from the experiments that 
 the \textsc{BlindGreedy} algorithm performs rather well, but also \textsc{BlindRandom}
algorithm reduces the amount of computed distances substantially, especially if we enforce having non-zero lower bounds
first. The data that we observed can serve as weak evidence
of the conjecture: the blind greedy algorithm computes
a spanner with $O(n)$ edges. The noteworthy fact is that quasi-sorted variants produce spanners which are much closer to the complete graph
(\textsc{BlindQuasiSortedGreedy} is worse, requiring all the edges). It would seem plausible that, if we have
access to approximate value of the distance, we could exploit this in the spanner construction, but we could not
find a working heuristics.

>>>>

\pgfplotsset{compat=1.14}

@@

However, we argue that in some situations, distance computations
in $\metricspace$ can be costly and $\complexity$ might be incomparable
with $n$. Our motivation comes from topological summaries
such as persistence diagrams~\cite{elz-topological} or Reeb graphs~\cite{reeb-survey}, which are of interest
in the field of topological data analysis. A persistence diagram
is a point set in $\R^2$, and the distance between two diagrams
is determined by a min-cost matching between the point sets.
If the diagrams have $N$ points, computing this matching requires
polynomial time in $N$, and $N$ might well be larger than $n$, the number
of diagrams considered (\cite{cohen2007stability}). For the case of Reeb graphs, the situation is even
worse: while several metrics on Reeb graphs have been proposed (\cite{bauer2014measuring}, \cite{de2016categorified},
\cite{di2016edit}),
not even an constant-factor approximation algorithm is known that runs
in polynomial time in the size of the graphs.
Another instance is a collection of high-resolution images
endowed with the Wasserstein (or Earth Movers) metric (\cite{rubner2000earth}).

@@

\myparagraph{Well-separated pair decomposition}
Given $t > 1$, two disjoint subsets $A, B$ of a metric space $(\metricspace, \dist)$ are called $t$-\textit{well-separated},
if 
\[
\forall a \in A \,\, \forall b \in B \,\, \dist(a, b) \geq t \max(\diam(A), \diam(B))
\]
A well-separated pair decomposition (WSPD) is a set of unordered pairs of sets $\{ \{A_1, B_1 \}, 
\dots, \{A_s, B_s\} \}$ such that each pair $\{A_i, B_i\}$ is $s$-well-separated, and for every unordered pair $\{a, b\}$ of distinct points of $\metricspace$ there exists a unique $j$ such that $a \in A_j$ and $b \in B_j$.
The notion of WSPD was introduced by Callahan and Kosaraju \cite{cal-kos-wspd} for Euclidean spaces.
Har-Peled and Mendel \cite{hm-fast} 
introduced the notion of net-trees and
generalized the results of \cite{cal-kos-wspd} for WSPD, proving the following:
\begin{enumerate}
    \item A net-tree for a metric space with $n$ points can be constructed in $2^{O(\mbox{dim})} n \log n$ expected time.
    \item If $\{ \{A_1, B_1 \}, 
\dots, \{A_s, B_s\} \}$ is an $\eps / 16$-WSPD on $\metricspace$, and $a_i \in A_i, b_i \in B_i$ for $i = 1\dots s$
are chosen arbitrarily, then we get an $\eps$-spanner by taking $s$ edges $(a_i, b_i)$.
    \item For $\eps \in (0, 1]$,  an $\eps$-WSPD of size $n \eps^{-O({\dim})}$ can be constructed in $2^{O({\dim})} n \log n + n \eps ^{-O({\dim})}$ expected time. The algorithm uses the net-tree structure.
\end{enumerate}
The algorithm of constructing a net-tree is complicated and not easy to implement. Beygelzimer
et al. \cite{cover-trees} introduced the notion of a cover tree, which is a simpler data structure 
than net-trees. We mention in passing that cover trees can also be used for building
a spanner (this can be proven with the same methods), and we use cover trees for
building WSPD spanners in one of our implementations.

@@

\textsc{BlindQuasiSortedGreedy} tries to mimic the greedy spanner and hence appears as a natural
choice at first sight. However, anticipating the experimental results, the heuristic yields very poor
results. The reason is that no pair acquires useful lower bounds when only short distance are queried
(the greedy spanner does not have this issue because it knows the distance and hence does not need
lower bounds).
Generally speaking, short distances are good for sharp upper bounds, whereas long distances are
useful for lower bounds. This motivates \textsc{BlindQuasiSortedShaker}
which alternates between short and long distances.

@@

\section{Experiments on spanners}
We run experiments on the points sampled from the low-dimensional Euclidean space to investigate
experimentally the performance of these heuristics. Clearly, for this metric space, our
cost model is not meaningful since distance comparisons are cheap;
but we picked this environment for controlled experiments.
In order to test the \textsc{BlindQuasiSorted} algorithms we multiply the true distance by a factor from $[1,2]$ chosen uniformly.

@@

Figure \ref{fig:spanner_sparseness}
shows the number of edges of the spanner for various variants of 
blind and non-blind spanner constructions.
Note that for all blind spanner variants, the number of computed distances
is equal to the spanner size, while for the non-blind greedy spanner,
this number is always $\binom{n}{2}$ and for WPSD it is lower bounded
by the size of the spanner.
We can see that, even though none of the blind spanners can produce
spanners of the same quality (i.e., sparse) as the standard greedy algorithm,
\textsc{BlindGreedy} and all variants of \textsc{BlindRandom}
perform significantly better than both variants of \textsc{BlindQuasiSorted}.
Figure \ref{fig:spanner_ratio} shows the ratio of the number of edges to the number of points.
The ideal behavior is demonstrated by the non-blind greedy spanner,
for which this ratio stays practically constant, confirming the linear growth.
None of the blind algorithms seems to have this property, but among them
the blind greedy spanner is the best one. If we assume that the number of edges
is proportional to $n^\alpha$, then we can try to estimate $\alpha$ by 
linear regression (after taking $\log$).  We give in the table \ref{tbl:regr_coeff_spanner}
the estimated exponents $\alpha$ for \textsc{BlindGreedy} and standard
greedy algorithms. Note that even for the greedy algorithm these estimated
exponents can be significantly larger than 1,
which is explained by the fact that the number of points
on which we computed spanners is not large enough to clearly see
the linear dependence.

@@

WSPD spanners performed poorly in our experiments
on non-clustered data, while the plots in the extensive experimental study \cite{farshi2009experimental}
show that WSPD spanners are very sparse, outperformed only by the greedy algorithm.
We implemented two versions of  WSPD: one for the Euclidean case,
using quadtrees and the algorithm from \cite{hp-book}, and WSPD for general metric
spaces with cover trees (using the base $\tau=1.3$). 
They both give similar results, and we can only conclude
that the advantage of WSPD shows up on larger point sets than the ones we deal with.
The paper \cite{farshi2009experimental} contains experiments for up to 30000 points,
and our blind algorithms, which have at least cubic complexity in the number of points,
are infeasible for such $n$.

We also tested higher dimensions and show the results for the best algorithm, \textsc{BlindGreedy},
in dimensions 2, 3 and 4 in the plot \ref{fig:blind_greedy_only}.
We can see that already in dimension 4, it produces a graph with roughly $\frac{1}{2}\binom{n}{2}$
edges for $700$ points, which clearly shows some degrading for higher dimensions. Still,
we remark that the WSPD spanner remains worse also in the higher-dimensional setup.

@@

Summing up, we can conclude from the experiments that 
 the \textsc{BlindGreedy} algorithm performs rather well, but also \textsc{BlindRandom}
algorithm reduces the amount of computed distances substantially, especially if we enforce having non-zero lower bounds
first. 
If the goal is to reduce the number of distance computations, these method seem to be more suitable than a WSPD spanner.
Since the linear spanner size of WSPDs does not show up in the experiments because of the relatively small values of $n$
tested, the experiments are not conclusive regarding the asymptotic size of the blind spanners.
Another noteworthy fact is that quasi-sorted variants produce spanners which are much closer to the complete graph
(\textsc{BlindQuasiSortedGreedy} is worse, requiring all the edges). It would seem plausible that, if we have
access to approximate value of the distance, we could exploit this in the spanner construction, but we could not
find a working heuristic.
